---
title: "adm_plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{adm_plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(admtools)
```

# Plotting age-depth models

This vignette describes the plotting options for plotting age-depth models.

## Simple age-depth models (class "adm")

As example, we use the age-depth model taken 4 km offshore in a carbonate platform modeled using CarboCAT Lite (Burgess 2013, 2023). This data is provided with the package.

```{r}
adm = tp_to_adm(t = CarboCATLite_data$time_myr,
                h = CarboCATLite_data$height_4_km_offshore_m,
                L_unit = "m",
                T_unit = "Myr")
```

The adm can be directly plotted using the standard `plot`comman from R. It can be modified using the arguments listed under `?plot.adm`

```{r}
plot(adm,
     lwd_destr = 1,
     lty_destr = 4,
     col_destr = "blue",
     lwd_acc = 1,
     lty_acc = 1,
     col_acc = "black")
```

Here the prefixes `lwd`, `lty`, and `col` stand for line width, line type, and color respectively. The suffix `destr` is for destructive intervals, the suffix `acc` for intervals with sediment accumulation.

### Adding axis labels

Axis lables can be added using the functions `T_axis_lab` (axis label of time axis) and `L_axis_lab` (length axis label). By default, they use the units associated with the age-depth model object. This behavior can be overwritten using the `label`and `unit` options. For more details, see the help page under `?T_axis_lab` and `L_axis_lab`. As example, we plot only accumulating parts, and add axis labels.

```{r}
plot(adm,
     lty_destr = 0,
     lwd_acc = 3)
T_axis_lab(label = "Elapsed Model Time")
L_axis_lab(label = "Stratigraphic Height")
```
## Multiple age-depth models (class "multiadm")


