---
title: "admtools: Getting Started"
output: rmarkdown::html_vignette
author: "Niklas Hohmann"
vignette: >
  %\VignetteIndexEntry{admtools: Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(admtools)
```

# Introduction

This vignette is an introduction to the *admtools* package.

## Installation

### From GitHub

To install the package from *GitHub*, first install the *remotes* package:

```{r, eval=FALSE}
install.packages("remotes")

```

Then run

```{r, eval=FALSE}
remotes::install_github(repo = "MindTheGap-ERC/admtools",
                        build_vignettes = TRUE,
                        ref = "HEAD",
                        dependencies = TRUE)
```

to install the latest version.

### From CRAN

To install the package from CRAN, run

```{r, eval=FALSE}
install.packages("admtools")
```

## Loading the package

Load the package using

```{r, eval=FALSE}
library(admtools)
```

This makes all functions in the package available.

## Getting help

Use

```{r, eval=FALSE}
help(package = "admtools")
```

to get an overview of the available help pages of the package, and

```{r, eval=FALSE}
?admtools
```

to view a simple help page for the package.

Vignettes are a long form of package documentation that provide more detailed examples. To list the available vignettes, use

```{r, eval=FALSE}
browseVignettes(package = "admtools") # opens in Browser
#or
vignette(package = "admtools")
```

## The adm and multiadm classes

The *admtools* package defines two main classes: `adm` and `multiadm`. The class `adm` represent a single age-depth model, from which information can be extracted (e.g. completeness, number of hiatuses, etc.) and that can be used to transform data between the stratigraphic domain and time domain. The `multiamd` class is a list of `adm` objects. `multiadm` objects are used to represent uncertainties of age-depth models.

## Conventions

In contrast to its name, the *admtools* package currently deals with time and height instead of age and depth. In this sense, the age-depth models are time-height models. Both time and height can be negative values. To handle ages, use time before the present. To handle depths, use height below a point of reference (e.g., the sediment surface).

## Example

This example explains the construction and application of `adm` objects. As example data we use outputs from CarboCAT Lite, a model of carbonate platform growth (Burgess 2013, 2023). This data is automatically loaded in the background by the package. To inspect it, you can load it into your workspace via

```{r}
data("CarboCATLite_data")
```

### Defining age-depth models

The standard constructor for age-depth models is `tp_to_adm` ("tiepoint to age-depth model"). It returns an objecto of class `adm`. This object combines information of stratigraphic heights and times and erosive interval. It allows to transform data between the stratigraphic and the time domain, and identify which data is destroyed due to hiatuses.

As example, I use the timing and stratigraphic positions of tie points taken from CarboCAT Lite to construct an age-depth model, and use the option to directly associate length and time units with it.

```{R}
# see ?tp_to_adm for detailed documentation
my_adm = tp_to_adm(t = CarboCATLite_data$time_myr,
                  h = CarboCATLite_data$height_2_km_offshore_m,
                  L_unit = "m",
                  T_unit = "Myr")
```

This age-depth model represents the relationship between elapsed model time and accumulated sediment 2 km offshore in a synthetic carbonate platform.

## Representation

Typing the name `my_adm` in the console will only tell that the generated variable is an age-depth model

```{r}
my_adm
```

To get a quick overview of the properties of `my_adm`, use `summary`:

```{R}
summary(my_adm)
```

If you want to inspect the insides of the object, use `str`:

```{r}
str(my_adm)
```

You can manually manipulate the fields of the `adm` object by treating it like a list. I do not recommend doing so, as it might result in unexpected behavior.

You can plot `adm` objects via the standard `plot` function. Here, I use the option to highlight hiatuses in red, and increase the linw width of the conservative ( = non-destructive) intervals.

```{R}
# see ?plot.adm for plotting options for adm objects
plot(my_adm,
     col_destr = "red",
     lwd_acc = 2)
T_axis_lab() # plot time axis label, see ?T_axis_lab for details
L_axis_lab() # plot height axis label, see ?L_axis_lab for details
```

### Extracting data from age-depth models

Use the functions `get_total_duration`, `get_total_thickness`, `get_completeness`, and `get_hiat_no` to extract information:

```{r}
get_total_duration(my_adm) #total time covered by the age-depth model
get_total_thickness(my_adm) # total thickness of section represented by the amd
get_completeness(my_adm) # stratigraphic completeness as proportion
get_incompleteness(my_adm) # stratigraphic incompleteness (= 1- strat. incompleteness)
get_hiat_no(my_adm) # number of hiatuses
```

For more detailed information, you can use

-   `get_hiat_duration` to get a vector of hiatus durations
-   `get_hiat_list` to get a list of hiatus positions and duration.

For example, to plot a histogram of hiatus durations, use

```{r}
hist(x = get_hiat_duration(my_adm),
     freq = TRUE,
     xlab = "Hiatus duration [Myr]",
     main = "Hiatus duration 2 km offshore")
```

The function `is_destructive` can be used to examine whether points in time coincide with hiatuses:

```{R}
is_destructive(my_adm,
               t = c(0.1,0.5)) 
```

### Transforming data between time and stratigraphic domain

#### Heights and times

The functions `get_height` and `get_time` are the workhorses to transform data using age-depth models.

-   `get_time` takes and adm object and vector of heights `h` (stratigraphic positions), and returns a vector of times
-   `get_height` takes an adm object and vector of times and returns a vector of associated positions

As example, say we want to know the time of deposition of the following stratigraphic positions:

```{r}
h = c(30,120) # stratigraphic positions
get_time(my_adm,
             h = h)
```

Conversely, to determine what parts of the section are deposited as a specific time, use

```{r}
t = c(0.2,1.4)
get_height(my_adm,
           t = t)
```

Here, the `NA` indicates that the time 1.4 coincides with erosion. If you want to know the stratigraphic position of the hiatus that coincides with that time, use the option `destructive = FALSE`:

```{r}
t = c(0.2,1.4)
get_height(my_adm,
           t = t,
           destructive = FALSE)
```

#### Phylogenetic trees

The `admtools` package can transform complex objects between the time and stratigraphic domain. This is done using the functions `strat_to_time` and `time_to_strat`.

As an example, we transform a chronogram (a phylogenetic tree where branch length represents time). For this, we first generate a tree following the birth-death model using the `ape` package.

```{r}
#install.packages("ape") Package for analyses of phylogenetics and evolution
# see ?ape::rlineage for help
set.seed(1)
tree_in_time = ape::rlineage(birth = 1.8,
                             death = 0.2,
                             Tmax = 2)
plot(tree_in_time) # see also ?ape::plot.phylo
axis(1)
mtext("Time [Myr]", side = 1, line = 2.5)
```

You can transform the tree using `time_to_strat`:

```{r}
tree_in_strat_domain = time_to_strat(obj = tree_in_time,
                                     x = my_adm)
```

Plotting the resulting tree along the tratigraphic column shows how the evolutinoary relationships would be preserved 2 km offshore in the simulated carbonate platform:

```{r}
plot(tree_in_strat_domain, direction = "upwards")
axis(side = 2)
mtext("Stratigraphic Height [m]",
      side = 2,
      line = 2)
```

#### Lists and time/stratigraphic series

*admtools* can transfrom lists from the time to the height domain and vice versa given they have elements with names `h` or `t`. These lists can be interpreted as time/stratigraphic series, where times and stratigraphic positions are associated with measured values. Note that these are not `t` objects as used by the `stats` package, as they will be generally heterodistant due to hte irregular nature of the age-depth transformation.

As example, we simulate trait evolution over 2 Myr using a Brownian motion, and transform the simulation into the stratigraphic domain.

```{r}
t = seq(0, 2, by = 0.001) # times
BM = function(t){
  #" Simulate Brownian motion at times t
  li = list("t" = t,
            trait_val = cumsum(c(0, rnorm(n = length(t) - 1, mean = 0, sd = sqrt(diff(t))))))
  return(li)
}
evo_list = BM(t)
plot(x = evo_list$t,
     y = evo_list$trait_val,
     xlab = "Time [Myr]",
     ylab = "Trait Value",
     type = "l")
```

```{r}
strat_list = time_to_strat(obj = evo_list,
                            x = my_adm)
plot(x = strat_list$h,
     y = strat_list$trait_val,
     type = "l",
     xlab = "Stratigraphic Height [m]",
     ylab = "Trait Value",
     main = "Trait Evolution 2 km Offshore")
```

Note the jump in traits generated by the erosional interval in `my_adm`.

## Further information

For information on estimating age-depth models from sedimentation rates, see

```{r, eval=FALSE}
vignette("adm_from_sedrate")
```

For information on estimating age-depth models from tracer contents of rocks and sediments, see

```{r, eval=FALSE}
vignette("adm_from_trace_cont")
```

## References

-   Burgess, Peter. "CarboCAT: A cellular automata model of heterogeneous carbonate strata." Computers & geosciences 53 (2013): 129-140. [DOI: 10.1016/j.cageo.2011.08.026](https://doi.org/10.1016/j.cageo.2011.08.026)
-   Burgess, Peter. "CarboCAT Lite v1.0.1". Zenodo 2023. [DOI: 10.5281/zenodo.8402578](https://doi.org/10.5281/zenodo.8402578)
