---
title: "admtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{admtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(admtools)
```

# Hello World

This vignette is an introduction to the _admtools_ adm package. As example data we use outputs from CarboCAT Lite, a model of carbonate platform growth. This data is provided with the package and loaded using the command below.

```{r}
load("data/CarboCatLite_data.rda")
```

## Defining age-depth models

The standard constructor for age-depth models is _tp_2_adm_ ("tiepoint to age-depth model"). It returns an _adm_ (age-depth model) object.
We use the timing and stratigraphic positions of tie points to construct an age-depth model, and use the option to directly associate length and time units with it.

```{R}
# see ?tp_2_adm for detailed documentation
my_adm = tp_2_adm(t = CarboCATLite_data$time_myr,
                  h = CarboCATLite_data$height_2_km_offshore_m,
                  L_unit = "m",
                  T_unit = "Myr")
```

You can use _summary_ to get a quick overview of the properties of the age-depth model
```{R}
summary(my_adm)
```

To get a visual representation, we plot it, and highlight hiatuses in red

```{R}
# see ?plot.adm for plotting options for adm objects
plot(my_adm,
     col_hiat = "red",
     lwd_cons = 3)
```

The function _is_destructive_ can be used to examine whether time points coincide with hiatuses:

```{R}
is_destructive(adm = my_adm,
               t = c(0.1,0.5)) 
```

## Transforming data between time and stratigraphic domain

The functions _get_height_ and _get_time_ are the workhorses to transform data using age-depth models.

* _get_time_ takes and adm object and vector of heights _h_  (stratigraphic positions), and returns a vector of times
* _get_height_ takes an adm object and vector of times and returns a vector of associated positions