---
title: "Structure and classes of the admtools package"
output: rmarkdown::html_vignette
author: "Niklas Hohmann"
vignette: >
  %\VignetteIndexEntry{Structure and classes of the admtools package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(admtools)
```

# Introduction

This vignette provides an overview of the larger scale structure of the `admtools` package and the classes used therein.

## S3 Classes

### S3 class `adm`

#### Description

The S3 class `adm` represents **a**ge **d**epth **m**odels. Structurally, they are lists with five fields:

-   `t` : numeric vector, time points

-   `h` : numeric vectors, heights

-   `destr`: logical vector, is the interval destructive

-   `T_unit` : `NULL` or a string, time unit

-   `L_unit` : `NULL` or a string, length unit

`h[i]` is the height at time `t[i]`. `h` and `t` must be of same length and have at least 2 elements, `t` must be strictly increasing and `h` must be nondecreasing. `length(destr)` must be identical to `length(t) - 1`. `destr[i] == TRUE` indicates that the time interval from `t[i]` to `t[i+1]` is destructive and no information is preserved. Whether tie points are considered destructive is determined by the function `is_destructive`. Geologically, `destr[i] == TRUE` should imply `h[i] == h[i+1]` , as no sediment is accumulated during hiatuses.

#### Construction

The following functions construct `adm` objects:

-   `tp_to_adm` for construction from tie points
-   `sac_to_adm` for construction from sediment accumulation curves `sac`
-   `split_multiadm` for extraction from multiple age-depth models `multiadm`

#### Logic

The following functions examine the logic of `adm` objects:

-   `is_adm` to check for validity of an adm object

#### Representation

The following functions yield a representation of `adm` objects:

-   `plot.adm` for plotting
-   `print.adm` for printing to the console
-   `summary.adm` to provide a quick summary of an object

#### Modification

The following functions modify `adm` objects:

-   `set_L_unit` and `set_T_unit` to change units

#### Extracting information

The following functions extract information from `adm` objects:

-   `condensation` and `condensation_fun` to extract condensation
-   `get_L_unit` and `get_T_unit` to extract data
-   `get_completeness` and `get_incompleteness`
-   `get_height`
-   `get_time`
-   `get_hiat_duration`
-   `get_hiat_list`
-   `get_hiat_no`
-   `get_hiat_pos`
-   `get_total_duration`
-   `get_total_thickness`
-   `is_destructive`
-   `sed_rate_l` and `sed_rate_l_fun` to extract sedimentation rate in height domain
-   `sed_rate_t` and `sed_rate_t_fun` to extract sedimentation rate in time domain
-   `max_height.adm` extracts highest stratigraphic point in adm
-   `min_height.adm` extracts lowest stratigraphic point in adm
-   `min_time.adm` extracts timing of first tie point in adm
-   `max_time.adm` extracts timing of last tie point in adm
-   `get_L_tp` and `get_T_tp` to extract tie points.

#### Transformation into other S3 classes

The following functions transform `adm` objects into other S3 classes:

-   `merge_adm_to_multiadm` into `multiamd`
-   `add_adm_to_multiadm` to add `adm` to `multiadm`
-   `anchor` to transform anchor `adm` at a tie point with uncertainty, resulting in a `multiadm`
-   `adm_to_ddc` to construct a depth-depth curve from two age-depth models

### S3 class `sac`

#### Description

The S3 class `sac` represents **s**ediment **a**ccumulation **c**urves. Structurally, they are lists with four fields:

-   `t` : numeric vector, time points
-   `h` : numeric vectors, heights
-   `T_unit` : `NULL` or a string, time unit
-   `L_unit` : `NULL` or a string, length unit

`h[i]` is the height at time `t[i]`. `h` and `t` must be of same length and have at least 2 elements, `t` must be increasing.

#### Construction

The following functions construct `sac` objects:

Standard constructor is `tp_to_sac` (tie point to sediment accumulation curve)

#### Logic

The following functions inspect the logic of `sac` objects:

-   `is_sac` to check validity of `sac` object

#### Representation

The following functions yield a representation of `sac` objects:

-   `plot.sac` for plotting
-   `print.sac` for printing to the console
-   `summary.sac` to provide a quick summary

#### Modification

The following functions modify `sac` objects:

-   `set_L_unit` and `set_T_unit` to change units

#### Extracting information

The following functions extract information from `sac` objects:

-   `get_L_unit` and `get_T_unit` to extract units
-   `get_T_tp` and `get_L_tp` to extract tie points
-   `get_total_duration` to extract total duration covered
-   `get_total_thickness`
-   `min_time` and `max_time`
-   `min_height` and `max_height`

#### Transformation into other S3 classes

The following functions transform `sac` objects into other S3 classes:

-   `sac_to_adm` to transform `sac` into the S3 class `adm`.

### S3 class `multiadm`

#### Description

The S3 class `multiadm` represents **multi**ple **a**ge **d**epth **m**odels.. Structurally, they are lists with the following elements:

-   `no_of_entries`: Positive integer, number of age depth models contained
-   `t` list of length `no_of_entries`. Each element contains a numeric vector
-   `h`: list of length `no_of_entries`. Each element contain a numeric vector
-   `destr`: list of length `no_of_entries`. Each element contain a logic
-   `T_unit` : `NULL` or a string, time unit
-   `L_unit` : `NULL` or a string, length unit

`h[[i]][j]` is the height of the i-th age-depth model at time `t[[i]][j]`. For each `i`, the quintuple `h[[i]]`, `t[[i]]`, `destr[[i]]`, `T_unit` and `L_unit` specify an `adm` object with the constraints as specified in the section *S3 class `adm`* (e.g., on monotonicity, length, etc.). `T_unit` and `L_unit` are shared among all age-depth models in an `multiamd` object.

#### Construction

The following functions construct `multiadm` objects:

-   `anchor` to construct `multiadm` from uncertain tie points and `adm` objects.
-   `merge_adm_to_multiadm` to construct `multiadm` from `adm` objects
-   `sedrate_to_multiadm` construct `multiadm` from info on sedimentation rates, see `vignette("adm_from_sedrate")`
-   `strat_cont_to_multiadm` construct `multiadm` from tracer information, see `vignette("adm_from_trace_cont")`

#### Logic

The following functions inspect the logic of `multiadm` objects:

-   `is_multiadm` to check for validity of multiadm object

#### Representation

The following functions yield a representation `multiadm` objects:

-   `plot.multiadm` for plotting
-   `print.multiadm` for printing to console
-   `summary.multiadm` for providing summary statistics

#### Modification

The following functions modify `multiadm` objects:

-   `merge_multiadm` to combine multiple `multiadm` objects
-   `set_L_unit` and `set_T_unit` to change units

#### Extracting information

The following functions extract information from `multiadm` objects:

-   `condensation` to extract condensation
-   `get_completeness` and `get_incompleteness`
-   `get_height`
-   `get_hiat_duration`
-   `get_hiat_list`
-   `get_hiat_no`
-   `get_hiat_pos`
-   `get_L_unit` and `get_T_unit` to extract data
-   `get_time`
-   `get_total_duration`
-   `get_total_thickness`
-   `is_destructive`
-   `sed_rate_l`
-   `sed_rate_t`
-   `get_T_tp` and `get_L_tp` to extract time and length/height tie points

#### Transformation into other S3 classes

The following functions transform `multiadm` objects into other S3 classes:

-   `split_multiadm` to split `multiadm` into list of `adm` objects
-   `mean_adm` , `median_adm` and `quantile_adm` to extract mean, median, and quantile age-depth model of `adm` class.m

### S3 class `ddc`

#### Description

The class `ddc` describes **d**epth-**d**epth **c**urves used for correlation between sections. They are lists with the following components:

- `h1`: numeric vector, heights in section 1
- `h2`: numeric vector, heights in section 2. Elements at the same position in `h1` and `h2` are coeval heights in the two sections
- `L_unit_1`: NULL or character, length unit of heights in section 1
- `L_unit_2`: NULL or character, length unit of heights in section 2
- `sec_1_name`: NULL or character, name of section 1
- `sec_2_name`: NULL or character, name of section 2

#### Construction

`ddc` can be constructed using

- `adm_to_ddc` to construct a depth-depth curve from two age-depth models
- `tp_to_ddc` to construct a depth-depth curve from coeval tie points

#### Logic

Not implemented yet

#### Representation

- `plot.ddc` for plotting depth-depth curves
- `summary.ddc` for a quick summary of the ddc object

#### Modification

Not implemented yet

#### Extracting information

Not implemented yet

#### Transformation inot other S3 classes

Not implemented yet



### S3 classes `stratlist` and `timelist`

`stratlist` and `timelist` inherit from the base `list`. They are list of stratigraphic positions or times, associated with other data (e.g. trait values, proxy values)

#### Description

-   `stratlist` is a list with one element named "h"
-   `timelist` is a list with one element named "t"

#### Construction

-   `stratlist` is returned by `time_to_strat.list`
-   `timelist` is returned by `strat_to_time.list`

#### Representation

-   `plot.stratlist` for plotting `stratlist`
-   `plot.timelist` for plotting `timelist`

## Methods implemented for external S3 classes

### S3 class `list`

`admtools` implements the following methods for `list`:

-   `strat_to_time.list`: Transform strat-val pairs into time domain
-   `time_to_strat.list`: Transform time-val pairs into strat domain

### S3 class `phylo`

`admtools` implements the following methods for `phylo`:

-   `strat_to_time.phylo`: Transform stratigraphic tree into time domain
-   `time_to_strat.phylo`: Transform time tree into strat domain

### S3 class `pre_paleoTS`

`admtools` implements the following methods for `pre_paleoTS` objects as defined by the `StratPal` package:

-   `strat_to_time.pre_paleoTS`: transform pre-paleoTS time series into time domain

-   `time_to_strat.pre_paleoTS`: transform pre-paleoTS time series into strat. domain

### Class `numeric`

-   `strat_to_time.numeric`: Transform vectors from stratigraphic domain to time domain. Wrapper around `get_time`
-   `time_to_strat.numeric`: Transform vectors from time to stratigraphic domain. Wrapper around `get_height`

## Plotting

The following functions are used for plotting:

-   `plot.adm` plotting for S3 class `adm`

-   `plot.multiadm` plotting for S3 class `multiadm`

-   `plot.sac` plotting for S3 class `sac`

-   `plot.timelist` for plotting `timelist`

-   `plot.stratlist` for plotting `stratlist`

-   `T_axis_lab` to annotate time axis

-   `L_axis_lab` to annotate length/depth axis

-   `plot_sed_rate_l` to plot sedimentation rate in length/depth domain

-   `plot_sed_rate_t` to plot sedimentation rate in time domain

-   `plot_condensation`

-   `plot_erosive_intervals` to highlight erosive intervals, called after `plot.adm`

### Internals

The following functions are used internally and not exposed to users. They can be accessed via `admtools:::function_name`.

-   `plot_destr_parts`

-   `plot_acc_parts`

-   `make_adm_canvas`

## Vignettes

-   `browseVignettes(package = "admtools")` to show a list of vignettes
