---
title: "Integration with the FossilSim and paleotree packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration with the FossilSim and paleotree packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(admtools)
```

This vignette explains the integration of `admtools` with the `FossilSim` and `paleotree` packages.

## Integration with the `FossilSim` package

`admtools` can transform phylogenetic trees coded as `phylo` objects as well as `fossils` and `taxonomy` objects as defined by the `FossilSim` package. This allows to transform phylogenetic trees with their associated taxonomic information and fossil locations between the time and the stratigraphic domain.

In contrast to `admtools`, `FossilSim` uses age (time before the present) instead of time. To successfully transform data generated by `FossilSim`, this must be reversed using `rev_dir`.

We give an example of this workflow

```{r}
set.seed(42)
# construct age-depth model
my_adm = tp_to_adm(t = CarboCATLite_data$time_myr,
                  h = CarboCATLite_data$height_2_km_offshore_m,
                  L_unit = "m",
                  T_unit = "Myr")
# simulate phylogenetic tree
t = ape::rbdtree(birth = 3, death = 1, Tmax = 2)
# simulate taxonomy along the tree
s = FossilSim::sim.taxonomy(tree = t)
# simulate fossils based on taxonomy
f = FossilSim::sim.fossils.poisson(rate = 4, taxonomy = s)
# plot tree with taxonomy and fossil times
FossilSim:::plot.fossils(f, tree = t, taxonomy = s, show.taxonomy = TRUE)

## transform everything into the strat domain
t_strat = time_to_strat(t, my_adm) # no transformation of time to age required
s_strat = s |>  # taxonomy object in the time domain
  rev_dir(ref = max_time(my_adm)) |> # convert age to time
  time_to_strat( my_adm, destructive = FALSE) |> # transform using age-depth model
  rev_dir(ref = max_height(my_adm)) # transform back into age
f_strat =  f |> # same here
  rev_dir(ref = max_time(my_adm)) |>
  time_to_strat( my_adm, destructive = TRUE)|> # destroy fossils coinciding with gaps
  rev_dir(ref = max_height(my_adm))

FossilSim:::plot.fossils(f_strat, tree = t_strat, taxonomy = s_strat, show.taxonomy = TRUE)
```

Important point are

-   `rev_dir` must be applied directly after data was generated and and directly before the data is further processed using `FossilSim`. As a principle, use `rev_dir` before data from FossilSim enters the `admtools` and `StratPal` ecosystem, and as it leaves the ecosystem.

-   By default, `FossilSim` plots have axis labels "Time before present", although the trees after the transformation can be trees in the stratigraphic domain. Because of this fixed reference point used, 0 in the above plot represents the uppermost height of the section, the axis units are lengths and not times, and the "Age" is stratigraphic position below the top of the section.

## Integration with `paleotree`

To combine the `admtools` package with the `paleotree` package, please convert the `paleotree` format into the `FossilSim` format as described in the `paleotree` vignette of the `FossilSim` vignette:

```{r eval=FALSE}
vignette("paleotree", package = "FossilSim")
```

After conversion you can proceed as described in the section *Integration with the `FossilSim` package*

## References

-   Barido-Sottani J, Pett W, O'Reilly JE, Warnock RCM. Fossilsim: An r package for simulating fossil occurrence data under mechanistic models of preservation and recovery. Methods Ecol Evol. 2019; 10: 835--840. <https://doi.org/10.1111/2041-210X.13170>
-   Bapst, D.W. (2012), paleotree: an R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution, 3: 803-807. <https://doi.org/10.1111/j.2041-210X.2012.00223.x>
-   Warnock R, Barido-Sottani J, Pett W, Joseph O, Stolz U (2024). *FossilSim: Simulation and Plots for Fossil and Taxonomy Data*. R package version 2.4.1, <https://CRAN.R-project.org/package=FossilSim>
